---
title: "Purrr02"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Relationship to base and plyr functions


### Why not base?

You need a way to iterate in R in a data-structure-informed way.

  * Iterate over elements of a list
  * Iterate over rows or columns of a 2-dimensional object  
  * Iterate over sub data frames induced by one or more factors
  * Iterate over tuples formed from the i-th element of several vectors of equal length

**All of this is absolutely possible with base R**, using `for()` loops or the "apply" functions, such as `apply()`, and `by()`.

Why might you do otherwise? 

  * they've never quite figured them out or been able to form the habit.

  * The user interface of the "apply"" functions is not as consistent as it could be, which slows down learning.
  
  * There's a tendency to return a vector (atomic or otherwise) or array, instead of data frame, with the original factor levels appearing in a names attribute.

### Why purrr?

purrr addresses "split-apply-combine":

  * The `map()` family of functions is highly internally consistent, making it easier to transfer expertise from one function to another.
  
  * Greater encouragement for type-safe simplification to atomic vector or data frame, producing output that is more ready for the next step.

  * Concise syntax for defining anonymous functions.


```{r}
library(purrr)
library(repurrrsive)
```


### `lapply()` vs. `purrr::map()`

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
lapply(got_chars[1:3],
       function(x) x[["name"]])
```
</div>
<div class = "col-md-6">
**purrr**
```{r}
map(got_chars[1:3], "name")
```
</div>
</div>


### `sapply()` vs. ¯\\\_(ツ)\_/¯

`sapply()` is a base function that attempts to apply a reasonable simplification to the output of `lapply()`. 
It's handy for interactive use, but due to the unpredictability of it return value.There is no equivalent in purrr.


```{r}
aliases1 <- sapply(got_chars[20:22], function(x) x[["aliases"]])
str(aliases1)
aliases2 <- sapply(got_chars[c(3, 22, 27)], function(x) x[["aliases"]])
str(aliases2)
```

With purrr, you would use `map()` to get a list back or `map_chr()` to get atomic character vector. If you use `map_chr()` when you should not, you'll get an informative error right away (shown below) and can adjust your approach accordingly.

```{r}
map_chr(got_chars[2:4], "aliases")
```

### `vapply()` vs. `map_*()`

Base `vapply()` requires you to specify a template for the return value and is described as a safer alternative to `sapply()`. 

The closest purrr functions are the type-specific mapping functions: `map_lgl()`, `map_int()`, `map_dbl()`, and `map_chr()` that are "list in, atomic vector out". 


<div class = "row">
<div class = "col-md-6">
**base**
```{r}
vapply(got_chars[1:3],
       function(x) x[["name"]],
       character(1))
```
</div>
<div class = "col-md-6">
**purrr**
```{r}
map_chr(got_chars[1:3], "name")
```
</div>
</div>

#### `vapply()` always simplifies

What's not to love with `vapply()` then? 

It suffers from the `drop = TRUE` vs `FALSE` problem we have when requesting a single row or column from a 2-dimensional object. Except `vapply()` has no `drop` argument to control this behavior. It's an example of the base functions being more difficult to program around. The template allows you to specify the form of each individual result, but there is no way to specify the form -- such as the dimension -- of the *overall* result.

I adapt this example from my real life, where I have `vapply()` inside a function and `n` is an argument to that function, i.e. it varies. Here I simply define `n` in the global environment prior to the `vapply()` call. Note how `vapply()` returns a 2 dimensional object in the first case and atomic vector in the second. 

As it says in the docs: "Simplification is always done in `vapply`." Believe it.

```{r}
f <- function(x, n) rep(x, n)
n <- 3
vapply(c("a", "b"), f, character(n), n = n)
n <- 1
vapply(c("a", "b"), f, character(n), n = n)
```

## ¯\\\_(ツ)\_/¯ vs. `map_df()`

The `purrr::map_df()` function is "list in, data frame out" and there is no true base equivalent. 

Given the centrality of data frames for analysis, it is handy to have a function to produce them, without resorting to `do.call()`.

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
l <- lapply(got_chars[23:25],
            `[`, c("name", "playedBy"))
mat <- do.call(rbind, l)
(df <- as.data.frame(mat, stringsAsFactors = FALSE))
```
</div>
<div class = "col-md-6">
**purrr**
```{r}
map_df(got_chars[23:25],
       `[`, c("name", "playedBy"))
```
</div>
</div>

The base workflow above gets trickier if you're extracting elements of disparate type. At that point, it may make more sense to use `vapply()` repeatedly. For comparability, we'll show similar using purrr's type-specific mapping, which is also safer than relying on automatic type conversion from `map_df()`.

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
data.frame(
  name = vapply(got_chars[23:25], `[[`,
                character(1), "name"),
  id = vapply(got_chars[23:25], `[[`,
              integer(1), "id"),
  stringsAsFactors = FALSE
)
```
</div>
<div class = "col-md-6">
**purrr**
```{r}
tibble::tibble(
  name = map_chr(got_chars[23:25], "name"),
  id = map_int(got_chars[23:25], "id")
)
```
</div>
</div>

### `mapply()` vs. `map2()`, `pmap()`

When you need to iterate over 2 or more vectors/lists in parallel, the base option is `mapply()`. Unlike the other apply functions, the first argument is `FUN`, the function to apply, and the multiple vector inputs are provided "loose" via `...`.

For exactly two vector inputs, purrr has `map2()`, with all the usual type-specific variants. For an arbitrary number of vector inputs, use purrr `pmap()` or type-specific variants, with the inputs packaged in a list. A very handy special case is when the input is a data frame, in which case `pmap_*()` applies `.f` to each row.

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
nms <- vapply(got_chars[16:18],
              `[[`, character(1), "name")
birth <- vapply(got_chars[16:18],
                `[[`, character(1), "born")
mapply(function(x, y) paste(x, "was born", y),
       nms, birth)
```
</div>
<div class = "col-md-6">
**purrr**
```{r}
nms <- got_chars[16:18] %>% 
  map_chr("name")
birth <- got_chars[16:18] %>% 
  map_chr("born")
map2_chr(nms, birth, ~ paste(.x, "was born", .y))

## and again, but with pmap()
df <- tibble::tibble(
  nms, 
  connector = "was born",
  birth
)
pmap_chr(df, paste)
```
</div>
</div>

### `aggregate()` vs. `dplyr::summarize()`

Consider a data frame, as opposed to a nested list. How do you split it into pieces, according to one or more factors, apply a function to the pieces, and combine the results?

```{r message = FALSE}
library(dplyr)
library(gapminder)
(mini_gap <- gapminder %>%
    filter(country %in% c("Canada", "Germany"), year > 2000) %>% 
    droplevels())
```

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
aggregate(lifeExp ~ country, mini_gap, mean)
```
</div>
<div class = "col-md-6">
**tidyverse**
```{r}
mini_gap %>% 
  group_by(country) %>% 
  summarize(lifeExp = mean(lifeExp))
```
</div>
</div>

Simple summaries of two variables for each country. We take the mean of life expectancy and of GDP per capita.

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
## formula method
aggregate(cbind(lifeExp, gdpPercap) ~ country, mini_gap, mean)

## data.frame method
aggregate(mini_gap[c("lifeExp", "gdpPercap")], list(mini_gap$country), mean)

## tapply() more general but output less useful here (data frame?)

## returns named vector
tapply(mini_gap$lifeExp, mini_gap$country, mean)

## returns list
tapply(mini_gap$lifeExp, mini_gap$country, mean, simplify = FALSE)
```
</div>
<div class = "col-md-6">
**tidyverse**
```{r}
mini_gap %>% 
  group_by(country) %>% 
  summarize_at(vars(lifeExp, gdpPercap), mean)
```
</div>
</div>

Bivariate summary of two variables for each country. We compute the correlation of life expectancy and year, for the full gapminder dataset now. On the base side, we can no longer use `aggregate()` or `tapply()` and need to graduate to `by()`.

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
## by() with simplification (the default)
by_obj <- by(gapminder, gapminder$country, function(df) cor(df$lifeExp, df$year))
head(by_obj)

## by() without simplification
by_obj <- by(gapminder, gapminder$country, function(df) cor(df$lifeExp, df$year),
   simplify = FALSE)
head(by_obj)
```
</div>
<div class = "col-md-6">
**tidyverse**
```{r}
gapminder %>% 
  group_by(country) %>% 
  summarize(cor = cor(lifeExp, year))
```
</div>
</div>


## `by()` vs. `tidyr::nest()`

Fit a linear model of life expectancy against year. On the tidyverse side, we now create a nested data frame, with one meta-row per country. Therefore we load tidyr to get `nest()`. The data needed for each country's linear model is stored as a list-column of country-specific data frame.

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
by_obj <- by(gapminder,
             gapminder$country, 
             function(df) lm(lifeExp ~ year, data = df))
str(by_obj[1:2], max.level = 1)
by_obj[[1]]
```
</div>
<div class = "col-md-6">
**tidyverse**
```{r}
library(tidyr)
library(purrr)
nested_df <- gapminder %>% 
  group_by(country, continent) %>% 
  nest() %>% 
  mutate(fit = map(data, ~ lm(lifeExp ~ year, data = .x)))
str(nested_df$fit[1:2], max.level = 1)
nested_df$fit[[1]]
```
</div>
</div>

What if you want to inspect the fits for Oceania? 

<div class = "row">
<div class = "col-md-6">
**base**
```{r}
o_countries <- as.character(unique(gapminder$country[gapminder$continent == "Oceania"]))
by_obj[names(by_obj) %in% o_countries]
```
</div>
<div class = "col-md-6">
**tidyverse**
```{r}
nested_df %>% 
  filter(continent == "Oceania") %>% 
  .$fit
```
</div>
</div>

Let's form a data frame with one row per country and variables for country, continent, estimated intercept, and estimated slope. 
<div class = "row">
<div class = "col-md-6">
**base**
```{r}
coefs <- lapply(by_obj, coef)
coefs <- do.call(rbind, coefs)
coefs <- data.frame(
  country = I(rownames(coefs)),
  coefs
)
coefs$continent <- gapminder$continent[match(coefs$country, gapminder$country)]
coefs$continent <- factor(coefs$continent, levels = levels(gapminder$continent))
coefs$country <- factor(coefs$country, levels = levels(gapminder$country))
head(coefs)
```
</div>
<div class = "col-md-6">
**tidyverse**
```{r}
nested_df %>% 
  mutate(coefs = map(fit, coef),
         intercept = map_dbl(coefs, 1),
         slope = map_dbl(coefs, 2)) %>% 
  select(country, continent, intercept, slope)
```
</div>
</div>




### Resources

[stat545, purrr, list-columns, nested data frames ](http://stat545.com/cm103_list-inspection-mapping.html)

[purrr tutorial](https://jennybc.github.io/purrr-tutorial/index.html)