---
title: "Purrr01"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Vectors and lists

### Atomic vectors

It is useful to understand lists as a data structure that generalizes atomic vectors. 

The garden variety R object is an **atomic vector** like these:

```{r}
(v_log <- c(TRUE, FALSE, FALSE, TRUE))
(v_int <- 1:4)
(v_doub <- 1:4 * 1.2)
(v_char <- letters[1:4])
```


Construct a vector by hand with the `c()` function. 
`1:n` shortcut.

To "index a vector" means to address specific elements or atoms, either for reading or writing. 

`x[something]`. 

there are several valid forms for `something`:

  * logical vector: keep elements of `x` for which `something` is `TRUE` and drop those for which it's `FALSE`
    ```{r}
    v_char[c(FALSE, FALSE, TRUE, TRUE)]
    v_char[v_log]
    ```

  * integer vector, all positive: the elements specified in `something` are kept

  * negative integers, all negative: the elements specified in `something` are dropped
    ```{r}
    v_doub[2:3]
    v_char[-4]
    ```

  * character vector: presumes that `x` is a named vector and the elements whose names are specified in `something` are kept *not shown here, since none of our vectors are named*


Do the exercises and you'll see it's possible to get an atomic vector of length zero and also to get elements that are `NA`. 

```{r}
v_int[0]
typeof(v_int[0])
v_doub[100]
typeof(v_doub[100])
```

Yes, there are different flavors of `NA`!

### Coercion

it's quite easy to convert them to another type. This is called **coercion**. 

There's a hierarchy of types: 

1. logical
2. integer
3. double
4. character

For explicit coercion, use the `as.*()` functions.

```{r}
v_log
as.integer(v_log)
v_int
as.numeric(v_int)
v_doub
as.character(v_doub)
as.character(as.numeric(as.integer(v_log)))
```

But coercion can also be triggered by other actions.

```{r}
v_doub_copy <- v_doub
str(v_doub_copy)
v_doub_copy[3] <- "uhoh"
str(v_doub_copy)
```

Our numeric vector was silently coerced to character. 

I end the discussion of atomic vectors with two specific examples of "being intentional about type".

  * Use of type-specific `NA`s when doing setup.
  * Use of `L` to explicitly request integer. 

```{r}
(big_plans <- rep(NA_integer_, 4))
str(big_plans)
big_plans[3] <- 5L
## note that big_plans is still integer!
str(big_plans)
## note that omitting L results in coercion of big_plans to double
big_plans[1] <- 10
str(big_plans)
```

### Lists

violates the constraints imposed by an atomic vector? 

  * Individual atoms might have length greater than 1.
  * Individual atoms might not be of the same flavor.
  
You need a list!

**A list is actually still a vector in R, but it's not an atomic vector.** 

We construct a list explicitly with `list()` but, like atomic vectors, most lists are created some other way in real life.

```{r}
(x <- list(1:3, c("four", "five")))
(y <- list(logical = TRUE, integer = 4L, double = 4 * 1.2, character = "character"))
(z <- list(letters[26:22], transcendental = c(pi, exp(1)), f = function(x) x^2))
```

list components can : 

  * Be heterogeneous
  * Have different lengths.
  * Have names. Or not. Or some of both.


### List indexing

There are 3 ways to index a list:

  1. With single square brackets. 
  
    * Note this **always returns a list**.
    
    ```{r}
    x[c(FALSE, TRUE)]
    y[2:3]
    z["transcendental"]
    ```

  1. With double square brackets. 
  
    * used to access a single component and it returns the "naked" component. You can request a component with a positive integer or by name.
    
    ```{r}
    x[[2]]
    y[["double"]]
    ```

  1. With the `$`, which you may already use to extract a single variable from a data frame. 
  
    ```{r}
    z$transcendental
    ```

### Vectorized operations

Consider how to square the integers 1 through `n`:

```{r}
n <- 5
res <- rep(NA_integer_, n) 
for (i in seq_len(n)) {
  res[i] <- i ^ 2
}
res
```

The R way is:

```{r}
n <- 5
seq_len(n) ^ 2
```

Element-wise or vectorized operations are "baked in" to a surprising degree in R. 

Here's a demo using `as.list()` to create the list version of an atomic vector.

```{r error = TRUE}
## elementwise exponentiation of numeric vector works
exp(v_doub)
## put the same numbers in a list and ... this no longer works :(
(l_doub <- as.list(v_doub))
exp(l_doub)
```

Use `purrr::map()`! The first argument is the list to operate on. The second is the function to apply.

```{r}
library(purrr)
map(l_doub, exp)
```

Vocabulary: we talk about this as "mapping the function `exp()` over the list `l_doub`". Conceptually, we loop over the elements of the list and apply a function.

```{r eval = FALSE}
my_list <- list(...)
my_output <- ## something of an appropriate size and flavor
for(i in seq_along(my_list)) {
  my_output[[i]] <- f(my_list([[i]]))
}
```

A major objective of this tutorial is to show you how to avoid writing these explicit `for()` loops yourself.



### Resources

[stat545, purrr, list-columns, nested data frames ](http://stat545.com/cm103_list-inspection-mapping.html)