---
title: "Logic in R - continue"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## The many flavors of R objects

### Vectors are everywhere

- R의 대부분의 object는 vector

- 하나의 숫자(scalar)도 길이가 1인 벡터

- __indexing__ : [ ](Square bracket)을 이용하면 해당 벡터로부터 각 요소를 분리하여 검사하거나 변환할 수 있음

    * R에서 indexing은 1부터 시작(다른 언어들은 0에서 시작)

```{r}
x <- 3 * 4
x
is.vector(x)
length(x)
x[2] <- 100
x
x[5] <- 3
x
x[11]
x[0]
```

- __vectorized__: R은 기본적으로 벡터단위로 연산 수행

    * 많은 연산함수(operations)들은 vectorization을 지원하기 위해 component-wise 연산을 수행
    
    * `for` loops 불필요

```{r}
x <- 1:4

## vectorized version:
(y <- x^2) 

## for loop version:
z <- vector(mode = mode(x), length = length(x))

for(i in seq_along(x)) {
  z[i] <- x[i]^2
}

identical(y, z)
```

vectorization 응용

```{r}
set.seed(1999)
rnorm(5, mean = 10^(1:5))
round(rnorm(5, sd = 10^(0:4)), 2)
```

vectorization을 편리하지만 위험하기도 하므로 항상 생성된 R object를 검사할 필함


__Recycling__

R은 vector를 recycle한다는 점에 주의

```{r}
(y <- 1:3)
(z <- 3:7)
y + z
(y <- 1:10)
(z <- 3:7)
y + z
```

__catenate function `c()`__

vector를 생성하거나 연결

```{r}
str(c("hello", "world"))
str(c(1:3, 100, 150))
```

__ type coercion__

- Plain vanilla R objects는 "atomic vectors"
- 포함된 모두 요소의 type이 동일해야함
- 아닌 경우 R이 "lowest common denominator" approach (usually character)를 적용하여 강제로 같도록 만듬


```{r}
(x <- c("cabbage", pi, TRUE, 4.3))
str(x)
length(x)
mode(x)
class(x)
```

atomic vector types: 

  * logical: TRUE's AND FALSE's, coerced into 1's and 0's
  * numeric: integers and double-precision float 
  * character

```{r}
n <- 8
set.seed(1)
(w <- round(rnorm(n), 2)) # numeric floating point
(x <- 1:n) # numeric integer
## another way to accomplish by hand is x <- c(1, 2, 3, 4, 5, 6, 7, 8)
(y <- LETTERS[1:n]) # character
(z <- runif(n) > 0.3) # logical
```

`str()`: inspect types of objects

    * `length()`, `mode()`, `class()`, `is.numeric()`, `is.logical()`, etc.
    
    
```{r}
str(w)
length(x)
is.logical(y)
as.numeric(z)
```

### Indexing a vector

Square brackets are used to index a vector. 

  * logical vector: keep elements associated with TRUEs, ditch the FALSEs
  * vector of positive integers specifying the keepers
  * vector of negative integers specifying the losers
  * character vector, naming the keepers

```{r}
w
names(w) <- letters[seq_along(w)]
w
w < 0
which(w < 0)
w[w < 0]
seq(from = 1, to = length(w), by = 2)
w[seq(from = 1, to = length(w), by = 2)]
w[-c(2, 5)]
w[c('c', 'a', 'f')]
```

### lists hold just about anything

- 리스트는 모든 type의 요소를 포함 가능

  * data.frames are lists.
  
  * many functions will return lists.(regression model)

- `list()` 

```{r}
## earlier: a <- c("cabbage", pi, TRUE, 4.3)
(a <- list("cabbage", pi, TRUE, 4.3))
str(a)
length(a)
mode(a)
class(a)
```

- List components는 names을 가짐

```{r}
names(a)
names(a) <- c("veg", "dessert", "myAim", "number")
a
a <- list(veg = "cabbage", dessert = pi, myAim = TRUE, number = 4.3)
names(a)
```

__Indexing a list__ 

- ["pepper shaker photos" in R for Data Science](http://r4ds.had.co.nz/vectors.html#lists-of-condiments) 


__more complicated list__

```{r}
(a <- list(veg = c("cabbage", "eggplant"),
           tNum = c(pi, exp(1), sqrt(2)),
           myAim = TRUE,
           joeNum = 2:6))
str(a)
length(a)
class(a)
mode(a)
```

Here's are ways to get a single list element:

```{r}
a[[2]] # index with a positive integer
a$myAim # use dollar sign and element name
str(a$myAim) # we get myAim itself, a length 1 logical vector
a[["tNum"]] # index with length 1 character vector 
str(a[["tNum"]]) # we get tNum itself, a length 3 numeric vector
iWantThis <- "joeNum" # indexing with length 1 character object
a[[iWantThis]] # we get joeNum itself, a length 5 integer vector
a[[c("joeNum", "veg")]] # does not work! can't get > 1 elements! see below
```


A case when one must use the double bracket approach, as opposed to the dollar sign, is when the indexing object itself is an R object; we show that above.

What if you want more than one element? You must index vector-style with single square brackets. Note that the return value will always be a list, unlike the return value from double square brackets, even if you only request 1 element.

```{r}
names(a)
a[c("tNum", "veg")] # indexing by length 2 character vector
str(a[c("tNum", "veg")]) # returns list of length 2
a["veg"] # indexing by length 1 character vector
str(a["veg"])# returns list of length 1
length(a["veg"]) # really, it does!
length(a["veg"][[1]]) # contrast with length of the veg vector itself
```


### Creating a data.frame explicitly

- `read.table()` : import data into data.frame 

- `data.frame()` : construct a data.frame.

```{r, tidy = FALSE}
n <- 8
set.seed(1)
(jDat <- data.frame(w = round(rnorm(n), 2),
                    x = 1:n,
                    y = I(LETTERS[1:n]),
                    z = runif(n) > 0.3,
                    v = rep(LETTERS[9:12], each = 2)))
str(jDat)
mode(jDat)
class(jDat)
```

- `I()`: protect the letters from being coerced to factor. We are ensuring we get a character vector. 

    * character-to-factor conversion happen in creating the $v$ variable above.
    
    * [Be the boss of your factors](http://www.stat.ubc.ca/~jenny/STAT545A/block08_bossYourFactors.html) 

- data.frames really are lists!

    * Double square brackets : to get individual variables.
    * Single square brackets : to get one or more variables, returned as a data.frame.
    * `subset(..., select = ...))` is how I would more likely do in a data analysis.

```{r, tidy = FALSE}
is.list(jDat) # data.frames are lists
jDat[[5]] # this works but I prefer ...
jDat$v # using dollar sign and name, when possible
jDat[c("x", "z")] # get multiple variables
str(jDat[c("x", "z")]) # returns a data.frame
identical(subset(jDat, select = c(x, z)), jDat[c("x", "z")])
```

- `as.data.frame()` : make a data.frame from a list.

```{r, tidy = FALSE}
## note difference in the printing of a list vs. a data.frame
(qDat <- list(w = round(rnorm(n), 2),
              x = 1:(n-1), ## <-- LOOK HERE! I MADE THIS VECTOR SHORTER!
              y = I(LETTERS[1:n])))
as.data.frame(qDat) ## does not work! elements don't have same length!
qDat$x <- 1:n ## fix the short variable x
(qDat <- as.data.frame(qDat)) ## we're back in business
```

### Indexing arrays, e.g. matrices

- matrix : a generalization of an atomic vector and the requirement that all the elements be of the same flavor still holds. 

    * General arrays are available in R, where a matrix is an important special case having dimension 2.


```{r}
## don't worry if the construction of this matrix confuses you; just focus on
## the product
jMat <- outer(as.character(1:4), as.character(1:4),
              function(x, y) {
                paste0('x', x, y)
                })
jMat
str(jMat)
class(jMat)
mode(jMat)
dim(jMat)
nrow(jMat)
ncol(jMat)
rownames(jMat)
rownames(jMat) <- paste0("row", seq_len(nrow(jMat)))
colnames(jMat) <- paste0("col", seq_len(ncol(jMat)))
dimnames(jMat) # also useful for assignment
jMat
```

- Indexing a matrix is very similar to indexing a vector or a list: use square brackets and index with logical, integer numeric (positive or negative), or character vectors. 

    * use a comma `,` to distinguish rows and columns.
    
    * The $i,j$-th element is the element at the intersection of row $i$ and column $j$ and is obtained with `jMat[i, j]`. 
    
    * Request an entire row or an entire column by simplying leaving the associated index empty. 
    
    * The `drop =` argument controls whether the return value should be an atomic vector (`drop = TRUE`) or a matrix with a single row or column (`drop = FALSE`). 
    
    
```{r}
jMat[2, 3]
jMat[2, ] # getting row 2
is.vector(jMat[2, ]) # we get row 2 as an atomic vector
jMat[ , 3, drop = FALSE] # getting column 3
dim(jMat[ , 3, drop = FALSE]) # we get column 3 as a 4 x 1 matrix
jMat[c("row1", "row4"), c("col2", "col3")]
jMat[-c(2, 3), c(TRUE, TRUE, FALSE, FALSE)] # wacky but possible
```

- R is a [column-major order](http://en.wikipedia.org/wiki/Row-major_order) language, in contrast to C and Python which use row-major order. 

```{r}
jMat[7]
jMat
```


```{r}
jMat[1, grepl("[24]", colnames(jMat))]
```

replacement values
```{r}
jMat["row1", 2:3] <- c("HEY!", "THIS IS NUTS!")
jMat
```

`%*%`: matrix multiplication.

__Creating arrays, e.g. matrices__

There are 3 main ways to create a matrix:

  * Filling a matrix with a vector
  * Glueing vectors together as rows or columns
  * Conversion of a data.frame


Fill a matrix with a vector.
```{r}
matrix(1:15, nrow = 5)
matrix("yo!", nrow = 3, ncol = 6)
matrix(c("yo!", "foo?"), nrow = 3, ncol = 6)
matrix(1:15, nrow = 5, byrow = TRUE)
matrix(1:15, nrow = 5,
       dimnames = list(paste0("row", 1:5),
                       paste0("col", 1:3)))
```

glueing vectors together.

```{r}
vec1 <- 5:1
vec2 <- 2^(1:5)
cbind(vec1, vec2)
rbind(vec1, vec2)
```

create a matrix from a data.frame.

```{r, tidy = FALSE}
vecDat <- data.frame(vec1 = 5:1,
                     vec2 = 2^(1:5))
str(vecDat)
vecMat <- as.matrix(vecDat)
str(vecMat)
```

Here we create a matrix from a data.frame, but experience the "silently convert everything to character" fail. 

```{r, tidy = FALSE}
multiDat <- data.frame(vec1 = 5:1,
                       vec2 = paste0("hi", 1:5))
str(multiDat)
(multiMat <- as.matrix(multiDat))
str(multiMat)
```

### Table of atomic R object flavors

+-----------+---------------+-----------+-----------+
| "flavor"  | type reported | mode()    | class()   |
|           | by typeof()   |           |           |
+===========+===============+===========+===========+
| character | character     | character | character |
+-----------+---------------+-----------+-----------+
| logical   | logical       | logical   | logical   |
+-----------+---------------+-----------+-----------+
| numeric   | integer       | numeric   | integer   |
|           | or double     |           | or double |
+-----------+---------------+-----------+-----------+
| factor    | integer       | numeric   | factor    |
+-----------+---------------+-----------+-----------+

> This should be legible no matter what.

```
+-----------+---------------+-----------+-----------+
| "flavor"  | type reported | mode()    | class()   |
|           | by typeof()   |           |           |
+===========+===============+===========+===========+
| character | character     | character | character |
+-----------+---------------+-----------+-----------+
| logical   | logical       | logical   | logical   |
+-----------+---------------+-----------+-----------+
| numeric   | integer       | numeric   | integer   |
|           | or double     |           | or double |
+-----------+---------------+-----------+-----------+
| factor    | integer       | numeric   | factor    |
+-----------+---------------+-----------+-----------+
```

## Acknowledgments

Above materials are derived in part from the following sources:

* Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)
* [R Language Definition](http://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html)
*[stat545](http://stat545.com/block004_basic-r-objects.html)
