---
title: "Be the boss of your factors"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Factors: where they fit in

Factors are the variable type that useRs love to hate. It is how we store truly categorical information in R. 

- The values a factor can take on are called the **levels**.

    * "Africa", "Americas", etc.
    * human-readable character strings, like "male/female" and "control/treated". 
    
- R is really storing integer codes 1, 2, 3, etc.

- stealth factor: The variable that you think of as character, but that is actually a factor (numeric!!). 

    * Base R has a burning desire to turn character information into factor. 
    
    * To shut this down, use `stringsAsFactors = FALSE` in `read.table()` and `data.frame()` or -- even better -- **use the tidyverse**! 


### The forcats package

forcats is a non-core package in the tidyverse. It is installed via `install.packages("tidyverse")`, but not loaded via `library(tidyverse)`. Load it yourself as needed via `library(forcats)`. 

    * Main functions start with `fct_`. 
    

### Load tidyverse, forcats, gapminder

```{r}
library(tidyverse)
library(forcats)
library(gapminder)
```

### Factor inspection

```{r}
str(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$continent)
```

To get a frequency table as a tibble, from a tibble, use `dplyr::count()`. 

    * To get similar from a free-range factor, use `forcats::fct_count()`.

```{r}
gapminder %>% 
  count(continent)
fct_count(gapminder$continent)
```

### Dropping unused levels

drop all the rows corresponding to a specific factor level

Watch what happens to the levels of `country` (= nothing) when we filter Gapminder to a handful of countries.

```{r}
nlevels(gapminder$country)
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries)
nlevels(h_gap$country)
```

Even though `h_gap` only has data for a handful of countries, we are still schlepping around all `r nlevels(gapminder$country)` levels from the original `gapminder` tibble.

How can you get rid of them? 

    * The base function `droplevels()` operates on all the factors in a data frame or on a single factor. 
    
    * The function `forcats::fct_drop()` operates on a factor.

```{r}
h_gap_dropped <- h_gap %>% 
  droplevels()
nlevels(h_gap_dropped$country)

## use forcats::fct_drop() on a free-range factor
h_gap$country %>%
  fct_drop() %>%
  levels()
```

Exercise: Filter the gapminder data down to rows where population is less than a quarter of a million, i.e. 250,000. Get rid of the unused factor levels for `country` and `continent` in different ways, such as:

  * `droplevels()`
  * `fct_drop()` inside `mutate()`
  * `fct_dopr()` with `mutate_at()` or `mutate_if()`

```{r, eval = FALSE, include = FALSE}
low_pop <- gapminder %>% 
  filter(pop < 250000)

low_pop %>%
  droplevels() %>% 
  group_by(country, continent) %>% 
  summarize(max_pop = max(pop))

low_pop %>%
  mutate(country = fct_drop(country),
         continent = fct_drop(continent)) %>% 
  group_by(country, continent) %>% 
  summarize(max_pop = max(pop))
```


### Change order of the levels, principled

By default, factor levels are ordered alphabetically.

How to order continent by frequency, esp. frequency barplots:

```{r}
## default order is alphabetical
gapminder$continent %>%
  levels()

## order by frequency
gapminder$continent %>% 
  fct_infreq() %>%
  levels()

## backwards!
gapminder$continent %>% 
  fct_infreq() %>%
  fct_rev() %>% 
  levels()
```

These two barcharts of frequency by continent differ only in the order of the continents. Which do you prefer?

```{r, fig.show = 'hold', out.width = '49%', echo = FALSE}
p <- ggplot(gapminder, aes(x = continent)) +
  geom_bar() +
  coord_flip()
p

gap_tmp <- gapminder %>% 
  mutate(continent = continent %>% fct_infreq() %>% fct_rev())

p <- ggplot(gap_tmp, aes(x = continent)) +
  geom_bar() +
  coord_flip()
p
```

order `country` by another variable:

```{r}
## order countries by median life expectancy
fct_reorder(gapminder$country, gapminder$lifeExp) %>% 
  levels() %>% head()

## order accoring to minimum life exp instead of median
fct_reorder(gapminder$country, gapminder$lifeExp, min) %>% 
  levels() %>% head()

## backwards!
fct_reorder(gapminder$country, gapminder$lifeExp, .desc = TRUE) %>% 
  levels() %>% head()
```

Example of why we reorder factor levels: often makes plots much better! 

```{r include = FALSE, eval = FALSE}
boxplot(Sepal.Width ~ Species, data = iris)
boxplot(Sepal.Width ~ fct_reorder(Species, Sepal.Width), data = iris)
boxplot(Sepal.Width ~ fct_reorder(Species, Sepal.Width, .desc = TRUE), data = iris)
```


```{r alpha-order-silly, fig.show = 'hold', out.width = '49%'}
gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia")
ggplot(gap_asia_2007, aes(x = lifeExp, y = country)) + geom_point()
ggplot(gap_asia_2007, aes(x = lifeExp, y = fct_reorder(country, lifeExp))) +
  geom_point()
```

Use `fct_reorder2()` when you have a line chart of a quantitative x against another quantitative y and your factor provides the color. 

    * This way the legend appears in some order as the data! 

```{r legends-made-for-humans, fig.show = 'hold', out.width = '49%'}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries) %>% 
  droplevels()
ggplot(h_gap, aes(x = year, y = lifeExp, color = country)) +
  geom_line()
ggplot(h_gap, aes(x = year, y = lifeExp,
                  color = fct_reorder2(country, year, lifeExp))) +
  geom_line() +
  labs(color = "country")
```

### Change order of the levels, "because I said so"

This resembles what we do when we move variables to the front with `dplyr::select(special_var, everything())`.

```{r}
h_gap$country %>% levels()
h_gap$country %>% fct_relevel("Romania", "Haiti") %>% levels()
```


### Recode the levels

Sometimes you have better ideas about what certain levels should be. This is called recoding.

```{r}
i_gap <- gapminder %>% 
  filter(country %in% c("United States", "Sweden", "Australia")) %>% 
  droplevels()

i_gap$country %>% levels()

i_gap$country %>%
  fct_recode("USA" = "United States", "Oz" = "Australia") %>% levels()
```

### Grow a factor

Let's create two data frames, each with data from two countries, dropping unused factor levels.

```{r}
df1 <- gapminder %>%
  filter(country %in% c("United States", "Mexico"), year > 2000) %>%
  droplevels()
df2 <- gapminder %>%
  filter(country %in% c("France", "Germany"), year > 2000) %>%
  droplevels()
```

The `country` factors in `df1` and `df2` have different levels.

```{r}
levels(df1$country)
levels(df2$country)
```

Can you just catenate them?

```{r}
c(df1$country, df2$country)
```

Umm, no. That is wrong on many levels! Use `fct_c()` to do this.

```{r}
fct_c(df1$country, df2$country)
```

Exercise: Explore how different forms of row binding work behave here, in terms of the `country` variable in the result.

```{r}
bind_rows(df1, df2)
rbind(df1, df2)
```


### References

[stat545, Be the boss of your factors](http://stat545.com/block029_factors.html)

