---
title: "Writing your own R functions"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Writing your own R functions (1)

### Load the Gapminder data

```{r}
library(gapminder)
str(gapminder)
```

### Max - min

### Get something that works

```{r}
## get to know the functions mentioned above
min(gapminder$lifeExp)
max(gapminder$lifeExp)
range(gapminder$lifeExp)

## some natural solutions
max(gapminder$lifeExp) - min(gapminder$lifeExp)
with(gapminder, max(lifeExp) - min(lifeExp))
range(gapminder$lifeExp)[2] - range(gapminder$lifeExp)[1]
with(gapminder, range(lifeExp)[2] - range(lifeExp)[1])
diff(range(gapminder$lifeExp))
```

### Turn the working interactive code into a function

```{r}
max_minus_min <- function(x) max(x) - min(x)
max_minus_min(gapminder$lifeExp)
```

### Test your function

__Test on new inputs__

```{r}
max_minus_min(1:10)
max_minus_min(runif(1000))
```

__Test on real data but *different* real data__

```{r}
max_minus_min(gapminder$gdpPercap)
max_minus_min(gapminder$pop)
```


__Test on weird stuff__

Now we try to break our function. Don't get truly diabolical (yet). Just make the kind of mistakes you can imagine making at 2am when, 3 years from now, you rediscover this useful function you wrote. Give your function inputs it's not expecting.

```{r}
max_minus_min(gapminder) ## hey sometimes things "just work" on data.frames!
max_minus_min(gapminder$country) ## factors are kind of like integer vectors, no?
max_minus_min("eggplants are purple") ## i have no excuse for this one
```

#### I will scare you now

Here are some great examples STAT545 students devised during class where the function __should break but it does not.__

```{r}
max_minus_min(gapminder[c('lifeExp', 'gdpPercap', 'pop')])
max_minus_min(c(TRUE, TRUE, FALSE, TRUE, TRUE))
```

### Check the validity of arguments

> Rule of Repair: When you must fail, fail noisily and as soon as possible.

#### stopifnot

`stopifnot()` is the entry level solution. I use it here to make sure the input `x` is a numeric vector.

```{r}
mmm <- function(x) {
  stopifnot(is.numeric(x))
  max(x) - min(x)
}
mmm(gapminder)
mmm(gapminder$country)
mmm("eggplants are purple")
mmm(gapminder[c('lifeExp', 'gdpPercap', 'pop')])
mmm(c(TRUE, TRUE, FALSE, TRUE, TRUE))
```

#### if then stop

`stopifnot()` doesn't provide a very good error message. The next approach is very widely used. Put your validity check inside an `if()` statement and call `stop()` yourself, with a custom error message, in the body.

```{r}
mmm2 <- function(x) {
  if(!is.numeric(x)) {
    stop('I am so sorry, but this function only works for numeric input!\n',
         'You have provided an object of class: ', class(x)[1])
  }
  max(x) - min(x)
}
mmm2(gapminder)
```

In addition to a gratuitous apology, the error raised also contains two more pieces of helpful info:

  * *Which* function threw the error.
  * Hints on how to fix things: expected class of input vs actual class.


### Wrap-up and what's next?

Here's the function we've written so far:

```{r}
mmm2
```


## Writing your own R functions (2)

```{r}
library(gapminder)
```

### Restore our max minus min function

```{r}
mmm <- function(x) {
  stopifnot(is.numeric(x))
  max(x) - min(x)
}
```

### Generalize our function to other quantiles

The max and the min are special cases of a __quantile__. Here are other special cases you may have heard of:

  * median = 0.5 quantile
  * 1st quartile = 0.25 quantile
  * 3rd quartile = 0.75 quantile
  
box plots

### Get something that works, again

The eventual inputs to our new function will be the data `x` and two probabilities.

First, play around with the `quantile()` function. 

```{r}
quantile(gapminder$lifeExp)
quantile(gapminder$lifeExp, probs = 0.5)
median(gapminder$lifeExp)
quantile(gapminder$lifeExp, probs = c(0.25, 0.75))
boxplot(gapminder$lifeExp, plot = FALSE)$stats
```

Now write a code snippet that takes the difference between two quantiles.

```{r}
the_probs <- c(0.25, 0.75)
the_quantiles <- quantile(gapminder$lifeExp, probs = the_probs)
max(the_quantiles) - min(the_quantiles)
```

### Turn the working interactive code into a function, again

```{r}
qdiff1 <- function(x, probs) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x = x, probs = probs)
  max(the_quantiles) - min(the_quantiles)
}
qdiff1(gapminder$lifeExp, probs = c(0.25, 0.75))
IQR(gapminder$lifeExp) # hey, we've reinvented IQR
qdiff1(gapminder$lifeExp, probs = c(0, 1))
mmm(gapminder$lifeExp)
```

### Argument names: freedom and conventions

I want you to understand the importance of argument names.

I can name my arguments almost anything I like. Proof:

```{r}
qdiff2 <- function(zeus, hera) {
  stopifnot(is.numeric(zeus))
  the_quantiles <- quantile(x = zeus, probs = hera)
  return(max(the_quantiles) - min(the_quantiles))
}
qdiff2(zeus = gapminder$lifeExp, hera = 0:1)
```

This is better:

```{r}
qdiff3 <- function(my_x, my_probs) {
 stopifnot(is.numeric(my_x))
  the_quantiles <- quantile(x = my_x, probs = my_probs)
  return(max(the_quantiles) - min(the_quantiles))
}
qdiff3(my_x = gapminder$lifeExp, my_probs = 0:1)
```


```{r}
qdiff1
```

### What a function returns

By default, a function returns the result of the last line of the body. I am just letting that happen with the line `max(the_quantiles) - min(the_quantiles)`. However, there is an explicit function for this: `return()`. I could just as easily make this the last line of my function's body:

```{r eval = FALSE}
return(max(the_quantiles) - min(the_quantiles))
```

You absolutely must use `return()` if you want to return early based on some condition, i.e. before execution gets to the last line of the body. Otherwise, you can decide your own conventions about when you use `return()` and when you don't.

### Default values: freedom to NOT specify the arguments

What happens if we call our function but neglect to specify the probabilities?

```{r}
qdiff1(gapminder$lifeExp)
```

Oops! At the moment, this causes a fatal error. It can be nice to provide some reasonable default values for certain arguments. In our case, it would be crazy to specify a default value for the primary input `x`, but very kind to specify a default for `probs`.

We started by focusing on the max and the min, so I think those make reasonable defaults. Here's how to specify that in a function definition.

```{r}
qdiff4 <- function(x, probs = c(0, 1)) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs)
  return(max(the_quantiles) - min(the_quantiles))
}
```

Again we check how the function works, in old examples and new, specifying the `probs` argument and not.

```{r}
qdiff4(gapminder$lifeExp)
mmm(gapminder$lifeExp)
qdiff4(gapminder$lifeExp, c(0.1, 0.9))
```

### Check the validity of arguments, again

EXERCISE FOR THE READER: upgrade our argument validity checks in light of the new argument `probs`

```{r}
## problems identified during class
## we're not checking that probs is numeric
## we're not checking that probs is length 2
## we're not checking that probs are in [0,1]
```

### Wrap-up and what's next?

Here's the function we've written so far:

```{r}
qdiff4
```


















## Writing your own R functions (3)

### Where are we going?

- tackle `NA`s
- the special argument `...` 
- formal testing

### Load the Gapminder data and Restore our max minus min function

```{r}
library(gapminder)

qdiff4 <- function(x, probs = c(0, 1)) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs)
  return(max(the_quantiles) - min(the_quantiles))
}
```

### Be proactive about `NA`s

Many built-in R functions have an `na.rm =` argument through which you can specify how you want to handle `NA`s. Typically the default value is `na.rm = FALSE` and typical default behavior is to either let `NA`s propagate or to raise an error. Let's see how `quantile()` handles `NA`s:

```{r}
z <- gapminder$lifeExp
z[3] <- NA
quantile(gapminder$lifeExp)
quantile(z)
quantile(z, na.rm = TRUE)
```

So `quantile()` simply will not operate in the presence of `NA`s unless `na.rm = TRUE`.

```{r}
qdiff5 <- function(x, probs = c(0, 1)) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs, na.rm = TRUE)
  return(max(the_quantiles) - min(the_quantiles))
}
qdiff5(gapminder$lifeExp)
qdiff5(z)
```

This works but it is dangerous. 

We could add an `na.rm =` argument to our own function. But at least we're giving the user a way to control the behavior around `NA`s.

```{r}
qdiff6 <- function(x, probs = c(0, 1), na.rm = TRUE) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs, na.rm = na.rm)
  return(max(the_quantiles) - min(the_quantiles))
}
qdiff6(gapminder$lifeExp)
qdiff6(z)
qdiff6(z, na.rm = FALSE)
```

### The useful but mysterious `...` argument

```{r}
qdiff7 <- function(x, probs = c(0, 1), na.rm = TRUE, ...) {
  the_quantiles <- quantile(x = x, probs = probs, na.rm = na.rm, ...)
  return(max(the_quantiles) - min(the_quantiles))
}
```


```{r}
set.seed(1234)
z <- rnorm(10)
quantile(z, type = 1)
quantile(z, type = 4)
all.equal(quantile(z, type = 1), quantile(z, type = 4))
```

Now we can call our function, requesting that quantiles be computed in different ways.

```{r}
qdiff7(z, probs = c(0.25, 0.75), type = 1)
qdiff7(z, probs = c(0.25, 0.75), type = 4)
```

The special argument `...` is very useful when you want the ability to pass arbitrary arguments down to another function, but without constantly expanding the formal arguments to your function. 

You will also encounter the `...` argument in many built-in functions -- read up [on `c()`](http://www.rdocumentation.org/packages/base/functions/c) or [`list()`](http://www.rdocumentation.org/packages/base/functions/list). 



### Use `testthat` for formal unit tests

The [`testthat` package](https://github.com/hadley/testthat) provides excellent facilities for this, with a distinct emphasis on automated unit testing of entire packages. 

We will construct a test with `test_that()` and, within it, we put one or more *expectations* that check actual against expected results. 

```{r}
library(testthat)
test_that('invalid args are detected', {
  expect_error(qdiff7("eggplants are purple"))
  expect_error(qdiff7(iris))
  })
test_that('NA handling works', {
  expect_error(qdiff7(c(1:5, NA), na.rm = FALSE))
  expect_equal(qdiff7(c(1:5, NA)), 4)
})
```

No news is good news! Let's see what test failure would look like. Let's revert to a version of our function that does no `NA` handling, then test for proper `NA` handling. We can watch it fail.

```{r}
qdiff_no_NA <- function(x, probs = c(0, 1)) {
  the_quantiles <- quantile(x = x, probs = probs)
  return(max(the_quantiles) - min(the_quantiles))
}
test_that('NA handling works', {
  expect_that(qdiff_no_NA(c(1:5, NA)), equals(4))
})
```



### Resources

Packages for runtime assertions (the last 3 seem to be under more active development than `assertthat`):

  * `assertthat` on [CRAN](https://cran.r-project.org/web/packages/asserthat/index.html) and [GitHub](https://github.com/hadley/assertthat) *the Hadleyverse option*
  * `ensurer` on [CRAN](https://cran.r-project.org/web/packages/ensurer/index.html) and [GitHub](https://github.com/smbache/ensurer) *general purpose, pipe-friendly*
  * `assertr` on [CRAN](https://cran.r-project.org/web/packages/assertr/index.html) and [GitHub](https://github.com/tonyfischetti/assertr) *explicitly data pipeline oriented*
  * `assertive` on [CRAN](https://cran.r-project.org/web/packages/assertive/index.html) and [Bitbucket](https://bitbucket.org/richierocks/assertive) *rich set of built-in functions*

Hadley Wickham's book [Advanced R](http://adv-r.had.co.nz)

  * Section on [defensive programming](http://adv-r.had.co.nz/Exceptions-Debugging.html#defensive-programming)


[stat545, Writing your own R functions](http://stat545.com/cm102_writing-functions.html)