---
title: "Purrr04 - advanced"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Specifying the function in map() + parallel mapping

### Load packages

```{r}
library(purrr)
library(repurrrsive)
```

### `map()` overview

Recall the usage of purrr's core `map()` function:

```{r eval = FALSE}
map(.x, .f, ...)
map(VECTOR_OR_LIST_INPUT, FUNCTION_TO_APPLY, OPTIONAL_OTHER_STUFF)
```

You can provide further arguments via `...`, but you don't have to. The above expands to something like this:

```{r eval = FALSE}
res <- vector(mode = "list", length = length(.x))
res[[1]] <- .f(.x[[1]], ...)
res[[2]] <- .f(.x[[2]], ...)
## and so on, until the end of .x
res
```

## `map()` function specification

One of the main reasons to use purrr is the flexible and concise syntax for specifying `.f`, the function to apply.

We demonstrate three more ways to specify general `.f`:

  * an existing function
  * an anonymous function, defined on-the-fly, as usual
  * a formula: this is unique to purrr and provides a very concise way to define an anonymous function

We work with the Game of Thrones character list, `got_chars`. Each character can have aliases, which are stored in a vector in each character's component. We pull out the aliases for three characters to use as our demo.

```{r}
aliases <- set_names(map(got_chars, "aliases"), map_chr(got_chars, "name"))
(aliases <- aliases[c("Theon Greyjoy", "Asha Greyjoy", "Brienne of Tarth")])
```

#### Existing function

```{r}
my_fun <- function(x) paste(x, collapse = " | ")
map(aliases, my_fun)
```

#### Anonymous function, conventional


```{r}
map(aliases, function(x) paste(x, collapse = " | ")) 
```

Alternatively you can simply name the function and provide `collapse` via `...`.

```{r}
map(aliases, paste, collapse = " | ")
```

#### Anonymous function, formula

We saved possibly the best for last.

purrr provides a very concise way to define an anonymous function: as a formula. This should start with the `~` symbol and then look like a typical top-level expression, as you might write in a script. Use `.x` to refer to the input, i.e. an individual element of the primary vector or list. 

```{r}
map(aliases, ~ paste(.x, collapse = " | "))
```

#### Workflow advice

A development process for the above might look like this:

```{r}
(a <- map(got_chars, "aliases")[[19]]) ## OOPS! NULL --> a useless example
(a <- map(got_chars, "aliases")[[16]]) ## ok good
paste(a, sep = " | ")                  ## OOPS! not what I want
paste(a, collapse = " | ")             ## ok good
got_chars[15:17] %>%                   ## I am a programming god
  map("aliases") %>% 
  map_chr(paste, collapse = " | ")
```

### List to data frame

The `enframe()` function from tibble takes a named vector and promotes the names to a proper variable.


```{r}
aliases <- set_names(map(got_chars, "aliases"), map_chr(got_chars, "name"))
map_chr(aliases[c(3, 10, 20, 24)], ~ paste(.x, collapse = " | ")) %>% 
  tibble::enframe(value = "aliases")
```

Alternative way to get same data frame

```{r}
tibble::tibble(
  name = map_chr(got_chars, "name"),
  aliases = got_chars %>% 
    map("aliases") %>% 
    map_chr(~ paste(.x, collapse = " | "))
) %>% 
  dplyr::slice(c(3, 10, 20, 24))
```


## Recap

These are the different ways to specify the function `.f` in the `map()`-type functions in purrr.

```{r eval = FALSE}
map(aliases, function(x) paste(x, collapse = "|")) 
map(aliases, paste, collapse = "|")
map(aliases, ~ paste(.x, collapse = " | "))
```

### Parallel map

#### `map2()`

What if you need to map a function over two vectors or lists in parallel?

You can use `map2()`:

```{r eval = FALSE}
map2(.x, .y, .f, ...)
map(INPUT_ONE, INPUT_TWO, FUNCTION_TO_APPLY, OPTIONAL_OTHER_STUFF)
```

`map2()` has all the type-specific friends you would expect: `map2_chr()`, `map2_lgl()`, etc.


First, obtain the two inputs.

```{r}
nms <- got_chars %>% 
  map_chr("name")
birth <- got_chars %>% 
  map_chr("born")
```

Now map over both with an existing function:

```{r}
my_fun <- function(x, y) paste(x, "was born", y)
map2_chr(nms, birth, my_fun) %>% head()
```

Anonymous function:

```{r}
map2_chr(nms, birth, function(x, y) paste(x, "was born", y)) %>% head()
```

Anonymous function via formula. Use `.x` and `.y` to refer to the individual elements of the two primary inputs.

```{r}
map2_chr(nms[16:18], birth[16:18], ~ paste(.x, "was born", .y)) %>% tail()
```


#### `pmap()`

What if you need to map a function over **two or more** vectors or lists in parallel?

You can use `pmap()`:

```{r eval = FALSE}
pmap(.l, .f, ...)
map(LIST_OF_INPUT_LISTS, FUNCTION_TO_APPLY, OPTIONAL_OTHER_STUFF)
```

*words*

```{r}
df <- got_chars %>% {
  tibble::tibble(
    name = map_chr(., "name"),
    aliases = map(., "aliases"),
    allegiances = map(., "allegiances")
  )
}
my_fun <- function(name, aliases, allegiances) {
  paste(name, "has", length(aliases), "aliases and",
        length(allegiances), "allegiances")
}
df %>% 
  pmap_chr(my_fun) %>% 
  tail()
```




## Resources

[stat545, purrr, list-columns, nested data frames ](http://stat545.com/cm103_list-inspection-mapping.html)

[purrr tutorial](https://jennybc.github.io/purrr-tutorial/index.html)


If you just wanted to solve this problem, you could let [`jsonlite`](https://github.com/jeroenooms/jsonlite) simplify the JSON for you. Other packages for list handling include [`listless`](https://bitbucket.org/graumannlabtools/listless), [`rlist`](https://renkun.me/rlist/).