---
title: "Purrr03 - core"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Explore the example lists: Wes Anderson, Game of Thrones, GitHub


### Load packages

Load purrr and repurrrsive, which contains recursive list examples.

```{r}
library(purrr)
library(repurrrsive)
```

List inspection is very important and also fairly miserable. Below I offer interactive widgets for our example lists. If you wish to do the same locally, install the [listviewer](https://CRAN.R-project.org/package=listviewer) package.

```{r}
# install.packages('listviewer')
library(listviewer)
```

### Inspect and explore

Before you can apply a function to every element of a list, you'd better understand the list!

You need to develop a toolkit for list inspection. Be on the look out for:

  * What is the length of the list?
  * Are the components homogeneous, i.e. do they have the same overall structure, albeit containing different data?
  * Note the length, names, and types of the constituent objects.


#### `str()`

`str()` can help with basic list inspection, although it's still rather frustrating. Learn to love the `max.level` and `list.len` arguments. You can use them to keep the output of `str()` down to a manageable volume.

Once you begin to suspect or trust that your list is homogeneous, i.e. consists of sub-lists with similar structure, it's often a good idea to do an in-depth study of a single element. In general, remember you can combine list inspection via `str(..., list.len = x, max.level = y)` with single `[` and double `[[` square bracket indexing.

The repurrrsive package provides examples of lists. We explore them below, to lay the groundwork for other lessons, and to demonstrate list inspection strategies.

#### listviewer

It's often more helpful to explore such lists with an interactive widget. If you installed the listviewer package, replicate this experience locally via `listviewer::jsonedit(got_chars, mode = "view")`.

### Wes Anderson color palettes

`wesanderson` is a simple list containing color palettes from the [wesanderson package](https://cran.r-project.org/package=wesanderson). Each component is a palette, named after a movie, and contains a character vector of colors as hexadecimal triplets.

```{r}
str(wesanderson)
```

#### listview `wesanderson`

```{r echo = FALSE}
jsonedit(wesanderson, mode = "view", elementId = "wesanderson")
```

### Game of Thrones POV characters

`got_chars` is a list with information on the `r length(got_chars)` point-of-view characters from the first five books in the Song of Ice and Fire series by George R. R. Martin. Retrieved from [An API Of Ice And Fire](https://anapioficeandfire.com). Each component corresponds to one character and contains `r length(got_chars[[1]])` components which are named atomic vectors of various lengths and types.

```{r}
str(got_chars, list.len = 3)
str(got_chars[[1]], list.len = 8)
```

#### listview `got_chars`

```{r echo = FALSE}
jsonedit(number_unnamed(got_chars), mode = "view", elementId = "got_chars")
```

### GitHub users and repositories

`gh_users` is a list with information on 6 GitHub users. `gh_repos` is a nested list, also of length 6, where each component is another list with information on up to 30 of that user's repositories. Retrieved from the [GitHub API](https://developer.github.com/v3/).

```{r}
str(gh_users, max.level = 1)
```

#### listview `gh_users`

```{r echo = FALSE}
jsonedit(number_unnamed(gh_users), mode = "view", elementId = "gh_users")
```

#### listview `gh_repos`

```{r echo = FALSE}
jsonedit(number_unnamed(gh_repos), mode = "view", elementId = "gh_repos")
```



## Introduction to map(): extract elements

### Load packages

Load purrr and repurrrsive, which contains recursive list examples. listviewer provides the interactive list viewing widgets.

```{r}
library(purrr)
library(repurrrsive)
library(listviewer)
```

### Vectorized and "list-ized" operations

This lesson picks up where [the primer on vectors and lists](https://jennybc.github.io/purrr-tutorial/bk00_vectors-and-lists.html#vectorized_operations) left off. 

Many operations "just work" in a vectorized fashion in R:

```{r}
(3:5) ^ 2
sqrt(c(9, 16, 25))
```

Someone -- but not you! -- has written a `for()` loop:

```
for (i in 1:n) {
  output[[i]] <- f(input[[i]])
}
```

Automatic vectorization is possible because our input is an atomic vector: always of length one, always of uniform type.

What if the input is a list? 

You have to be more intentional to apply a function `f()` to each element of a list, i.e. to "list-ize" computation. 

`purrr::map()` is a function for applying a function to each element of a list.(see `lapply()` in base R). 


```{r}
map(c(9, 16, 25), sqrt)
```

A template for basic `map()` usage:

```{r eval = FALSE}
map(YOUR_LIST, YOUR_FUNCTION)
```


#### Name and position shortcuts

We want the elements with name "name". purrr has a short cut to create a function that extracts an element based on its name.

```{r}
map(got_chars[1:4], "name")
```

A companion shortcut is used if you provide a positive integer to `map()`, based on element position.

```{r}
map(got_chars[5:8], 3)
```

To recap, here are two shortcuts for making the `.f` function that `map()` will apply:

  * provide "TEXT" to extract the element named "TEXT"
    - equivalent to `function(x) x[["TEXT"]]`
  * provide `i` to extract the `i`-th element
    - equivalent to `function(x) x[[i]]`

You will frequently see `map()` used together with the pipe `%>%`. These calls produce the same result as the above.

```{r eval = FALSE}
got_chars %>% 
  map("name")
got_chars %>% 
  map(3)
```


#### Type-specific map

`map()` always returns a list. But in that case, you might prefer a simpler object: **an atomic vector**. Tt is best to use a type-specific variant of `map()`.

This is more efficient than using `map()` to get a list and then simplifying the result in a second step. Also purrr will alert you to any problems, i.e. if one or more inputs has the wrong type or length. 


variants of `map()`, with the target type conveyed by the name:
    
    * `map_chr()``map_lgl()`, `map_int()`, `map_dbl()`

```{r}
map_chr(got_chars[9:12], "name")
map_chr(got_chars[13:16], 3)
```

  
#### Extract multiple values

What if you want to retrieve multiple elements? 

```{r}
got_chars[[3]][c("name", "culture", "gender", "born")]
```

We use single square bracket indexing and a character vector to index by name. 

    * To paraphrase Chambers, ["everything that happens in R is a function call"](http://adv-r.had.co.nz/Functions.html#all-calls) and indexing with `[` is no exception.

It feels (and maybe looks) weird, but we can map `[` just like any other function. Recall `map()` usage:

```{r eval = FALSE}
map(.x, .f, ...)
```

The function `.f` will be `[`. And we finally get to use `...`! This is where we pass the character vector of the names of our desired elements. We inspect the result for two characters.

```{r}
x <- map(got_chars, `[`, c("name", "culture", "gender", "born"))
str(x[16:17])
```

Some people find this ugly and might prefer the `extract()` function from magrittr.

```{r}
library(magrittr)
x <- map(got_chars, extract, c("name", "culture", "gender", "born"))
str(x[18:19])
```

#### Data frame output

How can we "stack up" these results row-wise, i.e. one row per user and variables for "name", "gender", etc.? 

    * A data frame would be the perfect data structure for this. This is what `map_df()` is for.

```{r}
map_df(got_chars, extract, c("name", "culture", "gender", "id", "born", "alive"))
```

Notice how the variables have been automatically type converted. It's a beautiful thing. 

When programming, it is safer to explicitly specify type and build your data frame the usual way.

```{r}
library(tibble)
got_chars %>% {
  tibble(
       name = map_chr(., "name"),
    culture = map_chr(., "culture"),
     gender = map_chr(., "gender"),       
         id = map_int(., "id"),
       born = map_chr(., "born"),
      alive = map_lgl(., "alive")
  )
}
```



## (Example)Simplifying data from a list of GitHub users


### Load packages

A great use of purrr's `map()` functions is to dig information out of a non-rectangular data structure(such as JSON or XML from an API) and create a neat data frame. 
If you are lucky it's JSON, which is less aggravating, and readily converts to a list you can work with in R.

Here we explore some lists obtained from the [GitHub API](https://developer.github.com/v3/). 

```{r message = FALSE}
library(repurrrsive)
library(listviewer)
library(jsonlite)
library(dplyr)
library(tibble)
library(purrr)
```

### Get several GitHub users 

The [repurrrsive package](https://github.com/jennybc/repurrrsive#readme) provides information on 6 GitHub users in an R list named `gh_users`.

`gh_users` is *a recursive list*:

  * one component for each of the 6 GitHub users
  * each component is, in turn, a list with info on the user

We have no clue what is in this list. 

Use `str()` with arguments such as `max.level` and `list.len`. 
```{r}
str(gh_users, max.level = 1)
str(gh_users[[1]], list.len = 6)
```

You can also use `listviewer::jsonedit()` to explore it interactively:

```{r echo = FALSE}
jsonedit(gh_users, mode = "view", elementId = "gh_users")
```


### Name and position shortcuts

Who are these GitHub users? We want the element with name "login", so we do this:

by name

```{r}
map(gh_users, "login")
```

by position

```{r}
map(gh_users, 18)
```

by pipe `%>%`

```{r eval = FALSE}
gh_users %>% 
  map("login")
gh_users %>% 
  map(18)
```

### Type-specific map

`map()` always returns a list. But you might prefer a simpler object: **an atomic vector**.

there are other variants of `map()`, with the target type conveyed by the name:

  * `map_chr()`, `map_lgl()`, `map_int()`, `map_dbl()`
  
```{r}
map_chr(gh_users, "login")
map_chr(gh_users, 18)
```


### Extract multiple values

What if you want to retrieve multiple elements?

```{r}
gh_users[[3]][c("name", "login", "id", "location")]
```

The function `.f` will be `[`.

```{r}
x <- map(gh_users, `[`, c("login", "name", "id", "location"))
str(x[1:2])
```

Using `extract()` function from magrittr.

```{r}
x <- map(gh_users, magrittr::extract, c("login", "name", "id", "location"))
str(x[3:4])
```

### Data frame output

How can we "stack up" these results row-wise, i.e. one row per user and variables for "login", "name", etc.? 

This is what `map_df()` is for.

```{r}
map_df(gh_users, `[`, c("login", "name", "id", "location"))
```

safer, but more cumbersome way

```{r}
gh_users %>% {
  tibble(
       login = map_chr(., "login"),
        name = map_chr(., "name"),
          id = map_int(., "id"),
    location = map_chr(., "location")
  )
}
```

### Vector input to extraction shortcuts

Instead of providing a single name or position, we use a vector:

  * the `j`-th element addresses the `j`-th level of the hierarchy
  
```{r}
gh_repos %>%
  map_chr(c(1, 3))
## TO DO? I would prefer a character example :( but gh_repos is unnamed atm
```

### List inside a data frame

Step 1: Put the `gh_repos` list into a data frame, along with identifying GitHub usernames. The care and feeding of lists inside a data frame -- "list-columns" -- is the subject of its own lesson (*yet to be written / linked*), so I ask you to simply accept that this can be done.

We use the answer to the previous exercise to grab the 6 usernames and set them as the names on the `gh_repos` list. Then we use `tibble::enframe()` to convert this named vector into a tibble with the names as one variable and the vector as the other. This is a generally useful setup technique.

```{r}
(unames <- map_chr(gh_repos, c(1, 4, 1)))
(udf <- gh_repos %>%
    set_names(unames) %>% 
    enframe("username", "gh_repos"))
```

Build confidence by doing something modest on the list-column of repos. This is your introduction to another powerful, general technique: `map()` inside `mutate()`. Note we are now bringing the data frame wrangling tools from dplyr and tidyr to bear.

```{r}
udf %>% 
  mutate(n_repos = map_int(gh_repos, length))
```

This shows that we know how to operate on a list-column inside a tibble.

Figure out how to do what we want for a single user, i.e. for a single element of `udf$gh_repos`. Walk before you run.

How far to we need to drill to get a single repo? How do we create "one row's worth" of data for this repo? How do we do that for all repos for a single user?

```{r}
## one_user is a list of repos for one user
one_user <- udf$gh_repos[[1]]
## one_user[[1]] is a list of info for one repo
one_repo <- one_user[[1]]
str(one_repo, max.level = 1, list.len = 5)
## a highly selective list of tibble-worthy info for one repo
one_repo[c("name", "fork", "open_issues")]
## make a data frame of that info for all a user's repos
map_df(one_user, `[`, c("name", "fork", "open_issues"))
## YYAAAASSSSSSS
```

Now we scale this up to all of our users. Yes, we use mutate to `map()` inside a `map()`.

```{r}
udf %>% 
  mutate(repo_info = gh_repos %>%
           map(. %>% map_df(`[`, c("name", "fork", "open_issues"))))
```

The user-specific tibbles about each user's repos are now sitting in the `repo_info`. How do we simplify this to a normal data frame that is free of list-columns? Remove the `gh_repos` variable, which has served its purpose and use `tidyr::unnest()`.

```{r}
(rdf <- udf %>% 
   mutate(
     repo_info = gh_repos %>%
       map(. %>% map_df(`[`, c("name", "fork", "open_issues")))
   ) %>% 
   select(-gh_repos) %>% 
   tidyr::unnest())
```

Let's do a little manipulation with dplyr to find some of the more interesting repos and get repos from each user in front of our eyeballs. I get rid of forks and show the 3 repos for each user that have the most open issues. (Remember we are only working with the first 30 repos for each user -- I had to remember my open issue situation is much more grim than this table suggests.)

```{r}
rdf %>% 
  filter(!fork) %>% 
  select(-fork) %>% 
  group_by(username) %>%
  arrange(username, desc(open_issues)) %>%
  slice(1:3)
```



## Resources

[stat545, purrr, list-columns, nested data frames ](http://stat545.com/cm103_list-inspection-mapping.html)

[purrr tutorial](https://jennybc.github.io/purrr-tutorial/index.html)


If you just wanted to solve this problem, you could let [`jsonlite`](https://github.com/jeroenooms/jsonlite) simplify the JSON for you. Other packages for list handling include [`listless`](https://bitbucket.org/graumannlabtools/listless), [`rlist`](https://renkun.me/rlist/).