---
title: "Getting data in and out of R"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## File I/O overview

### Read a file

- Load the tidyverse and forcats.

    * readr 
    * forcats : to reorder factor levels.

```{r}
library(tidyverse)
library(forcats)
library(gapminder)
```

- Locate the Gapminder data.

```{r}
# gap_tsv <- system.file("gapminder.tsv", package = "gapminder")
gap_tsv <- gapminder
```

- Bring rectangular data in

    * `readr::read_delim()`, `read_csv()`, `read_tsv()`

```{r}
# gapminder <- read_tsv(gap_tsv)
str(gap_tsv, give.attr = FALSE)
```

    * The most noticeable difference between the readr functions and base is that readr does NOT convert strings to factors by default. 

```{r}
gapminder <- gapminder %>%
  mutate(country = factor(country),
         continent = factor(continent))
str(gapminder)
```

### write to a file

- create a country-level summary of maximum life expectancy.

```{r}
gap_life_exp <- gapminder %>%
  group_by(country, continent) %>% 
  summarise(life_exp = max(lifeExp)) %>% 
  ungroup()
gap_life_exp
```

- Write rectangular data out

    * `write_delim()`, `write_csv()`

```{r}
write_csv(gap_life_exp, "gap_life_exp.csv")
```

Let's look at the first few lines of `gap_life_exp.csv`. If you're following along, you should be able to open this file or, in a shell, use `head` on it.

```{r echo = FALSE, comment = NA}
"gap_life_exp.csv" %>%
  readLines(n = 6) %>% 
  cat(sep = "\n")
```

- Reordering the levels of the country factor


```{r}
head(levels(gap_life_exp$country)) # alphabetical order
gap_life_exp <- gap_life_exp %>% 
  mutate(country = fct_reorder(country, life_exp))
head(levels(gap_life_exp$country)) # in increasing order of maximum life expectancy
head(gap_life_exp)
```

### `saveRDS()` and `readRDS()`

If you have a data frame AND you have exerted yourself to rationalize the factor levels, you have my blessing to save it to file in a way that will preserve this hard work upon re-import. Use `saveRDS()`.

```{r}
saveRDS(gap_life_exp, "gap_life_exp.rds")
```

- `saveRDS()` serializes an R object to a binary file. 

- It's a special purpose, limited use function that I use in specific situations.

- `readRDS()`. 

```{r error = TRUE}
rm(gap_life_exp)
gap_life_exp <- readRDS("gap_life_exp.rds")
gap_life_exp
```

- `saveRDS()` is also very handy for saving non-rectangular objects, like a fitted regression model.

- You will eventually hear about `save()` + `load()` and even `save.image()`. 

    *  These functions encourage unsafe practices, like storing multiple objects together and even entire workspaces. There are legitimate uses of these functions, but not in your typical data analysis.


__Retaining factor levels upon re-import__

Concrete demonstration of how non-alphabetical factor level order is lost with `write_delim()` / `read_delim()` workflows but maintained with `saveRDS()` / `readRDS()`.

```{r}
(country_levels <- tibble(original = head(levels(gap_life_exp$country))))
write_csv(gap_life_exp, "gap_life_exp.csv")
saveRDS(gap_life_exp, "gap_life_exp.rds")
rm(gap_life_exp)
head(gap_life_exp) # will cause error! proving gap_life_exp is really gone 
gap_via_csv <- read_csv("gap_life_exp.csv") %>% 
  mutate(country = factor(country))
gap_via_rds <- readRDS("gap_life_exp.rds")
country_levels <- country_levels %>% 
  mutate(via_csv = head(levels(gap_via_csv$country)),
         via_rds = head(levels(gap_via_rds$country)))
country_levels
```

### `dput()` and `dget()`

One last method of saving and restoring data deserves a mention: `dput()` and `dget()`. 

`dput()` offers this odd combination of features: it creates a plain text representation of an R object which still manages to be quite opaque. If you use the `file =` argument, `dput()` can write this representation to file but you won't be tempted to actually read that thing.

`dput()` creates an R-specific-but-not-binary representation. Let's try it out.

```{r}
## first restore gap_life_exp with our desired country factor level order
gap_life_exp <- readRDS("gap_life_exp.rds")
dput(gap_life_exp, "gap_life_exp-dput.txt")
```

Now let's look at the first few lines of the file `gap_life_exp-dput.txt`.

```{r echo = FALSE, comment = NA}
"gap_life_exp-dput.txt" %>% 
  readLines(n = 6) %>% 
  cat(sep = "\n")
```

The partner function `dget()` reads this representation back in.

```{r}
gap_life_exp_dget <- dget("gap_life_exp-dput.txt")
country_levels <- country_levels %>% 
  mutate(via_dput = head(levels(gap_life_exp_dget$country)))
country_levels
```

Note how the original, post-reordering country factor levels are restored using the `dput()` / `dget()` strategy.


### Clean up

```{r}
file.remove(list.files(pattern = "^gap_life_exp"))
```


### Invertibility

conflicts between rules:

  * Write to plain text files
  * Break analysis into pieces: the output of script `i` is an input for script `i + 1`
  * Be the boss of factors: order the levels in a meaningful, usually non-alphabetical way
  * Avoid duplication of code and data

So what do I do?: 

- save and re-load R-specific binary files.

- [a workflow where each analytical project is an R package](http://carlboettiger.info/2012/05/06/research-workflow.html)

- [Using YAML frontmatter with CSV](http://blog.datacite.org/using-yaml-frontmatter-with-csv/).



### References

[Data import](http://r4ds.had.co.nz/data-import.html) chapter of [R for Data Science](http://r4ds.had.co.nz) by Hadley Wickahm and Garrett Grolemund.

Nine simple ways to make it easier to (re)use your data by Ethan P White, Elita Baldridge, Zachary T. Brym, Kenneth J. Locey, Daniel J. McGlinn, Sarah R. Supp.

  * First appeared here: PeerJ PrePrints 1:e7v2 <http://dx.doi.org/10.7287/peerj.preprints.7v2>
  * Published here: Ideas in Ecology and Evolution 6(2): 1?10, 2013. doi:10.4033/iee.2013.6b.6.f <http://library.queensu.ca/ojs/index.php/IEE/article/view/4608>
  * Section 4 "Use Standard Data Formats" is especially good reading.
  
Tidy data by Hadley Wickham.

  * In the Journal of Statistical Software Vol 59 (2014), Issue 10, 10.18637/jss.v059.i10: <http://www.jstatsoft.org/article/view/v059i10>
  * PDF also available here: <http://vita.had.co.nz/papers/tidy-data.pdf>

Data Manipulation with R [available via SpringerLink](http://ezproxy.library.ubc.ca/login?url=http://link.springer.com.ezproxy.library.ubc.ca/book/10.1007/978-0-387-74731-6/page/1) by Phil Spector, Springer (2008) &#124; [author webpage](http://www.stat.berkeley.edu/%7Espector/) &#124; [GoogleBooks search](http://books.google.com/books?id=grfuq1twFe4C&lpg=PP1&dq=data%2520manipulation%2520spector&pg=PP1#v=onepage&q=&f=false)

  * See Chapter 2 ("Reading and Writing Data")

[stat545, Getting data in and out of R](http://stat545.com/block026_file-out-in.html)

