---
title: "Logic in R"
author: "Dookyung Kim"
date: "2018-07-11"
output: 
    html_document:
        css: note.css
mainfont: NanumGothic
---

```{r, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width=80
)

```

## Vector

R에서 생성되는 거의 대부분의 object(예: data)는 벡터(Vector)


#### Types of vectors


R언어는 두 가지 종류의 벡터클레스(vector classes)가 존재

* Vectors (atomic vectors) 

    - 원소의 타입(type)이 모두 동일한 벡터
    - 예: logical, numbers, character strings

* Lists (generic vectors)
  
    - 원소의 타입이 이질적인 벡터(even other lists)
    - a hierarchical/tree-like structure


#### Atomic Vectors


R has six atomic vector types: 

|  typeof    |   mode    | storage.mode 
|:-----------|:----------|:-------------
| logical    | logical   | logical      
| double     | numeric   | double       
| integer    | numeric   | integer      
| character  | character | character    
| complex    | complex   | complex      
| raw        | raw       | raw          


## Conditionals

#### Logical (boolean) operations

|  Operator  |  Operation    |  Vectorized? 
|:-----------|:--------------|:-------------
| <code>x &#124; y</code>    |  or           |   Yes        
| `x & y`    |  and          |   Yes        
| `!x`       |  not          |   Yes        
| <code>x &#124;&#124; y</code> |  or           |   No         
| `x && y`   |  and          |   No         
|`xor(x,y)`  |  exclusive or |   Yes        


#### Vectorized?

```{r}
x = c(TRUE,FALSE,TRUE)
y = c(FALSE,TRUE,TRUE)
```

```{r}
x | y
x || y
```

```{r}
x & y
x && y
```


#### Length coercion

```{r}
x = c(TRUE,FALSE,TRUE)
y = c(TRUE)
z = c(FALSE,TRUE)
```

```{r}
x | y
y | z
```

```{r}
x & y
y & z
```

```{r}
x | z
```

#### Comparisons

  Operator  |  Comparison                |  Vectorized?
:-----------|:---------------------------|:-----------------
 `x < y`    |  less than                 |  Yes
 `x > y`    |  greater than              |  Yes
 `x <= y`   |  less than or equal to     |  Yes
 `x >= y`   |  greater than or equal to  |  Yes
 `x != y`   |  not equal to              |  Yes
 `x == y`   |  equal to                  |  Yes
 `x %in% y` |  contains                  |  Yes (for `x`)


```{r}
x = c("A","B","C")
z = c("A")
```

```{r}
x == z
x != z
x > z
```

```{r}
x %in% z
z %in% x
```

## Conditional Control Flow

`if` statements. 

*Note that `if` statements are **not** vectorized.*

```{r}
x = c(3,1)

if (3 %in% x)
  "Here!"
if (x >= 2)
  "Now Here!"
```

#### Collapsing logical vectors

There are a couple of helper functions for collapsing a logical vector down to a single value: `any`, `all`

```{r}
x = c(3,4)
```

```{r}
any(x >= 2)
all(x >= 2)
```

```{r}
!any(x >= 2)
if (any(x >= 2))
    print("Now There!")
```

#### Nesting Conditionals

```{r}
x = 3
if (x < 0) {
  "Negative"
} else if (x > 0) {
  "Positive"
} else {
  "Zero"
}
```

```{r}
x = 0
if (x < 0) {
  "Negative"
} else if (x > 0) {
  "Positive"
} else {
  "Zero"
}
```


## Error Checking


#### `stop` and `stopifnot`

```{r error=TRUE}
ok = FALSE
if (!ok)
  stop("Things are not ok.")

stopifnot(ok)
```


#### Style choices

```{r eval=FALSE}
# Do stuff
if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
} else if (condition_error) {
  stop("Condition error occured")
}
```

```{r eval=FALSE}
# Do stuff better
if (condition_error) {
  stop("Condition error occured")
}

if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
}
```


## Loops

#### `for` loops

```{r}
for(x in 1:10)
{
  cat(x^2,"")
}
```

```{r}
for(y in list(1:3, LETTERS[1:7], c(TRUE,FALSE)))
{
  cat(length(y),"")
}
```


#### `while` loops

Repeat until the given condition is **not** met (i.e. evaluates to `FALSE`)

```{r}
i = 1
res = rep(NA,10)

while (i <= 10) {
  res[i] = i^2
  i = i+1
}

res
```


#### `repeat` loops

Repeat until `break`

```{r}
i = 1
res = rep(NA,10)

repeat {
  res[i] = i^2
  i = i+1
  if (i > 10)
    break
}

res
```


#### Special keywords - `break` and `next`

These are special actions that only work *inside* of a loop

* `break` - ends the current *loop* (inner-most)
* `next` - ends the current *iteration*

```{r}
for(i in 1:10) {
    if (i %% 2 == 0)
        break
    cat(i,"")
}
```

```{r}
for(i in 1:10) {
    if (i %% 2 == 0)
        next
    cat(i,"")
}
```


#### Some helper functions

Often we want to use a loop across the indexes of an object and not the elements themselves. There are several useful functions to help you do this: `:`, `length`, `seq`, `seq_along`, `seq_len`, etc.

```{r}
4:7
length(4:7)
seq(4,7,by=1)
seq_along(4:7)
seq_len(length(4:7))
```


## Functions

#### When to use functions

The goal of a function should be to encapsulate a *small* *reusable* piece of code.

* Name should make it clear what the function does (think in terms of simple verbs).

* Functionality should be simple enough to be quickly understood.

* The smaller and more modular the code the easier it will be to reuse elsewhere.

* Better to change code in one location than code everywhere.


#### Function Parts

The two parts of a function are the arguments (`formals`) and the code (`body`).

```{r}
gcd = function(long1, lat1, long2, lat2) {
  R = 6371 # Earth mean radius in km
  # distance in km
  acos(sin(lat1)*sin(lat2) + cos(lat1)*cos(lat2) * cos(long2-long1)) * R
}

formals(gcd)
body(gcd)
```

#### Return values

There are two approaches to returning values: explicit and implicit return values.

*Explicit* - includes one or more `return` statements

```{r}
f = function(x) {
  return(x*x)
}
```


*Implicit* - value of the last statement is returned.

```{r}
f = function(x) {
  x*x
}
```

#### Returning multiple values

If we want a function to return more than one value we can group things using either vectors or lists.

```{r}
f = function(x) {
  c(x, x^2, x^3)
}

f(2)
f(2:3)
```


#### Argument names

When defining a function we are also implicitly defining names for the arguments, when calling the function we can use these names to pass arguments in a different order.


```{r}
f = function(x,y,z) {
  paste0("x=",x," y=",y," z=",z)
}
```

```{r,error=TRUE}
f(1,2,3)
f(z=1,x=2,y=3)
```

.column[
```{r,error=TRUE}
f(y=2,1,3)
f(y=2,1,x=3)
```

```{r,error=TRUE}
f(1,2,3,m=1)
```

#### Argument defaults

It is also possible to give function arguments default values so that they don't need to be provided every time the function is called.

```{r error=TRUE}
f = function(x,y=1,z=1) {
  paste0("x=",x," y=",y," z=",z)
}
```

```{r error=TRUE}
f()
f(x=3)
f(y=2,2)
```


#### Scope

R has generous scoping rules, if it can't find a variable in the functions body, it will look for it in the next higher scope, and so on.

```{r}
y = 1
f = function(x) {
  x+y
}
f(3)
```

```{r}
g = function(x) {
  y=2
  x+y
}
g(3)
```

#### 
Additionally, variables defined within a scope only persist for the duration of that scope, and do not overwrite variables at a higher scopes (unless you use the global assignment operator `<<-`, *which you shouldn't*)

```{r}
x = 1
y = 1
z = 1
f = function()
{
    y = 2
    g = function()
    {
      z = 3
      return(x + y + z)
    }
    return(g())
}
f()
c(x,y,z)
```


## Lazy evaluation

Arguments to R functions are lazily evaluated - meaning they are not evaluated until they are used 

```{r, error=TRUE}
f = function(x)
{
  cat("Hello world!\n")
  x
}

f(stop())
```


## Everything is a function

```{r}
`+`
typeof(`+`)
x = 4:1
`+`(x,2)
```

## Getting Help

Prefixing any function name with a `?` will open the related help file for that function.

```{r, eval=FALSE}
?`+`
?sum
```

For functions not in the base package, you can generally see their implementation by entering the function name without parentheses (or using the `body` function).

```{r}
lm
```

### Less Helpful Examples

```{r}
list

`[`

sum

`+`
```



## Acknowledgments

Above materials are derived in part from the following sources:

* Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)
* [R Language Definition](http://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html)

